Informing IPCC accounting of forest carbon using the global forest carbon database (ForC v4.0)

### 2022-04-15 doc/manuscript/ESSDmanuscript.Rmd
update variable mapping figure

@@

### Soil Organic Matter

Soil organic matter is defined as "Includes organic carbon in mineral and organic soils (including peat) to a specified depth chosen by the country and
applied consistently through the time series. Live fine roots (of less than the suggested diameter limit for belowground biomass) are included with soil organic matter where they cannot be distinguished from it empirically." [@ipcc_good_2003;@ipcc_2006_2006].

```{r fig_variable_mapping, out.width = "15cm", echo = FALSE, fig.cap = "\\textbf{Schematic illustrating the carbon pools quantified under IPCC accounting; ForC variables corresponding to the stock, increment, influx and outflux; and relationships among them.} In many cases, the match of ForC variables to IPCC criteria depends upon measurement protocols (e.g., min DBH). Additional caveats are as follows: 1- often excludes branch fall, assumes that change in foliage biomass is negligible (see note 7); 2- excludes large branch fall; also, under IPCC definitions, outflux from aboveground biomass should include all sizes, influx to deadwood should include only above the minimum diameter chosen for dead wood; 3- excludes belowground components;  4-excludes breakage into pieces less than dead wood threshold size; 5- may exclude woody mortality of stems <10 cm DBH and branch fall, decomposition of dead wood into sizes classified as litter; 6- measurements often limited to decomposition of relatively fine litter and may exclude branches and stems below the dead wood size threshold and/or the more decomposed layers of the O horizon; 7 - foliage production is generally measured by collecting leaf-fall, a method that assumes that the influx = outflux (foliage biomass is roughly constant year-to-year); 8 - excludes branch fall, which is necessary for a full accounting of woody productivity but is typically assumed negligible for calculations of net biomass change."}
#this is for a two-column figure. for one-column figure, use out.width = "8.3cm"
knitr::include_graphics("figures_tables/C_variable_mapping.png")
```

## Land use categories
IPCC defines land-use categories to include six categories -- Forest Land, Grassland, Wetlands, Cropland, Settlements, and Other Land [@ipcc_2006_2006].
Sub-divisions include land that has remained in a particular category for >20 years (e.g., Forest Land remaining Forest Land) and land that has been converted from one category to another in the past 20 years (e.g., Cropland converted to Forest Land). 
Definitions of forest are allowed to vary by country, but must be applied consistently.

Managed land...

@@

```{r fig_dominant_veg, out.width = "15cm", echo = FALSE, fig.cap = "\\textbf{Histograms of number of records in ForC relevant to (grey) and transferred to (black) EFDB, organized by (a) dominant vegetation type, (b) FAO ecozone, (c) continent, and (d) stand age.} Codes are as follows: .... "}
#this is for a two-column figure. for one-column figure, use out.width = "8.3cm"
knitr::include_graphics("figures_tables/Histogram_n_Relevant_and_Transferred_Records.png")
```

@@

library(knitr)
library(kableExtra)
knitr::kable(C_variables_table, format = "latex",
             longtable = TRUE,
             caption = "\\textbf{Numbers of records of ForC variables relevant to, and sent to, EFDB.} Relationships of variables to IPCC-defined forest C pools (Table 1) and to each other are illustrated in Figure 1. ") %>%
  kableExtra::kable_styling(full_width = TRUE) %>%
  row_spec(0, bold = TRUE)  %>%
  row_spec(1, bold = TRUE)  %>% #biomass
  row_spec(6, bold = TRUE)  %>% #agb
  row_spec(14, bold = TRUE) %>% #root
  row_spec(25, bold = TRUE)  %>% #dead wood
  row_spec(33, bold = TRUE)  %>% #litter
  row_spec(40, bold = TRUE)  %>% #som
  row_spec(44, bold = TRUE) #total
```

>>>>

```{r fig_variable_mapping, out.width = "14cm", echo = FALSE, fig.cap = "\\textbf{Schematic illustrating the carbon pools quantified under IPCC accounting; ForC variables corresponding to the stock, increment, influx and outflux; and relationships among them.} Correspondence of ForC variables to IPCC criteria often depends upon measurement protocols (e.g., min DBH). Additional caveats are as follows: (a,b) branch fall and mortality of stems below census min DBH, which are necessary for a full accounting of dead organic matter production but typically assumed negligible for calculations of biomass change, are excluded by common measurement practice (a) or ForC variable definition (b); (c) assumes that leaf production equals leaf fall, and that changes in foliage biomass are negligble; (d,e) belowground components excluded by common measurement practice (d) or ForC variable definition (e); (f) excludes movement of dead wood into litter through breakage or size reduction; (g) measurements often limited to litter horizon (OL) and may exclude larger branches and stems classified as litter and/or the more decomposed layers of the O horizon."}
#this is for a two-column figure. for one-column figure, use out.width = "8.3cm"
knitr::include_graphics("figures_tables/C_variable_mapping.png")
```

@@

### Soil Organic Matter/ Carbon

Soil organic matter/ carbon (SOM/ SOC), which *(statistic on how much C is typial)* [@ref], is defined by IPCC as "organic carbon in mineral and organic soils (including peat) to a specified depth chosen by the country and applied consistently through the time series" [@ipcc_good_2003;@ipcc_2006_2006]. 
Live fine roots (suggested diameter cutoff of 2 mm) may be included with soil organic matter when it is not feasible to distinguish them empirically.
The greatest source of methodological variation in measuring SOM/ SOC is sampling depth, which has a suggested default of 30 cm but may vary by country provided that consistent criteria are applied.

## Land use categories
IPCC defines land-use categories to include six categories -- Forest Land, Grassland, Wetlands, Cropland, Settlements, and Other Land [@ipcc_2006_2006].
Sub-divisions include land that has remained in a particular category for >20 years (e.g., Forest Land remaining Forest Land) and land that has been converted from one category to another in the past 20 years (e.g., Cropland converted to Forest Land). 
Forest Land
Definitions of forest are allowed to vary by country, but must be applied consistently.

Inventories are conducted only for managed land, which is defined by IPCC to include "all forests subject to some kind of human interactions (notably commercial management, harvest of industrial round-wood (logs) and fuelwood, production and use of wood commodities, and forest managed for amenity value or environmental protection if specified by the country), with defined geographical boundaries" [@ipcc_good_2003.

@@

```{r fig_dominant_veg, out.width = "15cm", echo = FALSE, fig.cap = "\\textbf{Histograms of number of records in ForC relevant to (grey) and transferred to (black) EFDB, organized by (a) dominant vegetation type, (b) FAO ecozone, (c) continent, and (d) stand age.} For dominant vegetation (a), 'Other' includes deciduous needleleaf, mixed broadleaf- needleleaf, non-woody vegetation (e.g., early successional), and incompletely classified or mixed forest types. For FAO ecozones (b), codes are as listed in the caption of Figure 2."}
#this is for a two-column figure. for one-column figure, use out.width = "8.3cm"
knitr::include_graphics("figures_tables/Histogram_n_Relevant_and_Transferred_Records.png")
```

@@

library(knitr)
library(kableExtra)
knitr::kable(C_variables_table, format = "latex",
             longtable = TRUE,
             caption = "\\textbf{Numbers of records of ForC variables relevant to, and sent to, EFDB.} Relationships of variables to IPCC-defined forest C pools (Table 1) and to each other are illustrated in Figure 1. ") %>%
  kableExtra::kable_styling(full_width = TRUE) %>%
  row_spec(0, bold = TRUE)  %>%
  row_spec(1, bold = TRUE)  %>% #biomass
  row_spec(6, bold = TRUE)  %>% #agb
  row_spec(14, bold = TRUE) %>% #root
  row_spec(25, bold = TRUE)  %>% #dead wood
  row_spec(33, bold = TRUE)  %>% #litter
  row_spec(40, bold = TRUE)  %>% #som
  row_spec(44, bold = TRUE)  #total

```

### 2022-04-15 doc/manuscript/ESSDmanuscript.tex
update variable mapping figure

\caption{\label{tab:table_pools}\textbf{IPCC-defined forest carbon pools with definitions and measurement methods.} Definitions from IPCC Table 1.1. (See Table 1.1 in IPCC guidance).}
\centering
\begin{tabu} to \linewidth {>{\raggedright}X>{\raggedright}X>{\raggedright}X>{\raggedright}X}
\hline
\textbf{pool} & \textbf{definition} & \textbf{important sources of estimate variation} & \textbf{IPCC guidance}\\
\hline
aboveground biomass & all biomass of living vegetation & min dbh & may exclude understory if minor component\\
\hline
 &  & include non-dicot trees? & ?\\
\hline
 &  & include dead standing? & no\\
\hline
 &  & biomass allometry & ?\\
\hline
belowground biomass & all biomass of live roots & all factors relevant to aboveground biomass & see above\\
\hline
 &  & allometry or assumed ratio of below- to above-ground biomass (R) & can estimate based on R\\
\hline
 &  & min root diameter & may exclude fine roots\\
\hline
 &  & sampling depth & ?\\
\hline
dead wood & all non-living woody biomass above a specified diameter, aboveground or belowground & min diameter & 10 cm default, but may be chosen by country\\
\hline
 &  & include belowground? & \vphantom{1} yes\\
\hline
litter & all non-living biomass, within specified size limits, in various states of decomposition above or within the mineral or organic soil & max diameter (= min diameter for deadwood) & 10 cm default, but may be chosen by country\\
\hline
 &  & min size (= size limit for soil organic matter) & ?\\
\hline
 &  & layers included & entire O horizon: litter (OL),  fumic (OF),  and  humic (OH) layers\\
\hline
 &  & include belowground? & yes\\
\hline
soil organic matter & organic carbon in mineral soils to a specified depth & sampling depth & 30 cm default, but may be chosen by country\\
\hline
\end{tabu}
\end{table}

@@

\subsubsection{Soil Organic Matter}

Soil organic matter is defined as ``Includes organic carbon in mineral
and organic soils (including peat) to a specified depth chosen by the
country and applied consistently through the time series. Live fine
roots (of less than the suggested diameter limit for belowground
biomass) are included with soil organic matter where they cannot be
distinguished from it empirically.''
\citep{ipcc_good_2003, ipcc_2006_2006}.

\begin{figure}
\includegraphics[width=15cm]{figures_tables/C_variable_mapping} \caption{\textbf{Schematic illustrating the carbon pools quantified under IPCC accounting; ForC variables corresponding to the stock, increment, influx and outflux; and relationships among them.} In many cases, the match of ForC variables to IPCC criteria depends upon measurement protocols (e.g., min DBH). Additional caveats are as follows: 1- often excludes branch fall, assumes that change in foliage biomass is negligible (see note 7); 2- excludes large branch fall; also, under IPCC definitions, outflux from aboveground biomass should include all sizes, influx to deadwood should include only above the minimum diameter chosen for dead wood; 3- excludes belowground components;  4-excludes breakage into pieces less than dead wood threshold size; 5- may exclude woody mortality of stems <10 cm DBH and branch fall, decomposition of dead wood into sizes classified as litter; 6- measurements often limited to decomposition of relatively fine litter and may exclude branches and stems below the dead wood size threshold and/or the more decomposed layers of the O horizon; 7 - foliage production is generally measured by collecting leaf-fall, a method that assumes that the influx = outflux (foliage biomass is roughly constant year-to-year); 8 - excludes branch fall, which is necessary for a full accounting of woody productivity but is typically assumed negligible for calculations of net biomass change.}\label{fig:fig_variable_mapping}
\end{figure}

@@

IPCC defines land-use categories to include six categories -- Forest
Land, Grassland, Wetlands, Cropland, Settlements, and Other Land
\citep{ipcc_2006_2006}. Sub-divisions include land that has remained in
a particular category for \textgreater20 years (e.g., Forest Land
remaining Forest Land) and land that has been converted from one
category to another in the past 20 years (e.g., Cropland converted to
Forest Land). Definitions of forest are allowed to vary by country, but
must be applied consistently.

Managed land\ldots{}

@@

\begin{figure}
\includegraphics[width=15cm]{figures_tables/Histogram_n_Relevant_and_Transferred_Records} \caption{\textbf{Histograms of number of records in ForC relevant to (grey) and transferred to (black) EFDB, organized by (a) dominant vegetation type, (b) FAO ecozone, (c) continent, and (d) stand age.} Codes are as follows: .... }\label{fig:fig_dominant_veg}
\end{figure}

>>>>

\caption{\label{tab:table_pools}\textbf{IPCC-defined forest carbon pools with definitions and measurement methods.} Definitions from IPCC Table 1.1. (See Table 1.1 in IPCC guidance).}
\centering
\begin{tabu} to \linewidth {>{\raggedright}X>{\raggedright}X>{\raggedright}X>{\raggedright}X}
\hline
\textbf{pool} & \textbf{definition} & \textbf{important sources of estimate variation} & \textbf{IPCC guidance}\\
\hline
aboveground biomass & all biomass of living vegetation & min dbh & may exclude understory if minor component\\
\hline
 &  & include non-dicot trees? & ?\\
\hline
 &  & include dead standing? & no\\
\hline
 &  & biomass allometry & ?\\
\hline
belowground biomass & all biomass of live roots & all factors relevant to aboveground biomass & see above\\
\hline
 &  & allometry or assumed ratio of below- to above-ground biomass (R) & can estimate based on R\\
\hline
 &  & min root diameter & may exclude fine roots\\
\hline
 &  & sampling depth & ?\\
\hline
dead wood & all non-living woody biomass above a specified diameter, aboveground or belowground & min diameter & 10 cm default, but may be chosen by country\\
\hline
 &  & include belowground? & \vphantom{1} yes\\
\hline
litter & all non-living biomass smaller than dead wood but larger than soil organic matter, in various states of decomposition both above or within the mineral or organic soil & max diameter (= min diameter for deadwood) & 10 cm default, but may be chosen by country\\
\hline
 &  & min size (= size limit for soil organic matter) & ?\\
\hline
 &  & layers included & entire O horizon: litter (OL),  fumic (OF),  and  humic (OH) layers\\
\hline
 &  & include belowground? & yes\\
\hline
soil organic matter & organic carbon in mineral soils to a specified depth & sampling depth & 30 cm default, but may be chosen by country\\
\hline
\end{tabu}
\end{table}

\begin{figure}
\includegraphics[width=14cm]{figures_tables/C_variable_mapping} \caption{\textbf{Schematic illustrating the carbon pools quantified under IPCC accounting; ForC variables corresponding to the stock, increment, influx and outflux; and relationships among them.} Correspondence of ForC variables to IPCC criteria often depends upon measurement protocols (e.g., min DBH). Additional caveats are as follows: (a,b) branch fall and mortality of stems below census min DBH, which are necessary for a full accounting of dead organic matter production but typically assumed negligible for calculations of biomass change, are excluded by common measurement practice (a) or ForC variable definition (b); (c) assumes that leaf production equals leaf fall, and that changes in foliage biomass are negligble; (d,e) belowground components excluded by common measurement practice (d) or ForC variable definition (e); (f) excludes movement of dead wood into litter through breakage or size reduction; (g) measurements often limited to litter horizon (OL) and may exclude larger branches and stems classified as litter and/or the more decomposed layers of the O horizon.}\label{fig:fig_variable_mapping}
\end{figure}

@@

\subsubsection{Soil Organic Matter/ Carbon}

Soil organic matter/ carbon (SOM/ SOC), which \emph{(statistic on how
much C is typial)} \citep{ref}, is defined by IPCC as ``organic carbon
in mineral and organic soils (including peat) to a specified depth
chosen by the country and applied consistently through the time series''
\citep{ipcc_good_2003, ipcc_2006_2006}. Live fine roots (suggested
diameter cutoff of 2 mm) may be included with soil organic matter when
it is not feasible to distinguish them empirically. The greatest source
of methodological variation in measuring SOM/ SOC is sampling depth,
which has a suggested default of 30 cm but may vary by country provided
that consistent criteria are applied.

@@

IPCC defines land-use categories to include six categories -- Forest
Land, Grassland, Wetlands, Cropland, Settlements, and Other Land
\citep{ipcc_2006_2006}. Sub-divisions include land that has remained in
a particular category for \textgreater20 years (e.g., Forest Land
remaining Forest Land) and land that has been converted from one
category to another in the past 20 years (e.g., Cropland converted to
Forest Land). Forest Land Definitions of forest are allowed to vary by
country, but must be applied consistently.

Inventories are conducted only for managed land, which is defined by
IPCC to include ``all forests subject to some kind of human interactions
(notably commercial management, harvest of industrial round-wood (logs)
and fuelwood, production and use of wood commodities, and forest managed
for amenity value or environmental protection if specified by the
country), with defined geographical boundaries''
{[}\citet{ipcc_good_2003}.

@@

\begin{figure}
\includegraphics[width=15cm]{figures_tables/Histogram_n_Relevant_and_Transferred_Records} \caption{\textbf{Histograms of number of records in ForC relevant to (grey) and transferred to (black) EFDB, organized by (a) dominant vegetation type, (b) FAO ecozone, (c) continent, and (d) stand age.} For dominant vegetation (a), 'Other' includes deciduous needleleaf, mixed broadleaf- needleleaf, non-woody vegetation (e.g., early successional), and incompletely classified or mixed forest types. For FAO ecozones (b), codes are as listed in the caption of Figure 2.}\label{fig:fig_dominant_veg}
\end{figure}
